Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-23 HRUIofMovingObjects/jrnl_article.tex
Changed symbol for resolution.

The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\Delta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
\end{align}
where the oversampling factor $p$ is applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$-factor of two has been used throughout this paper. The oversampling factor $q$ will be used to oversample prior to Capon beamforming and is one if not stated otherwise.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references). The maximal steering angle is approximately given by
\begin{align}
\theta_{max} = \alpha p \Delta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\Delta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\Delta$.
It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\Delta/2$ to $\Delta/2$. Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements. This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \Delta/2$.

>>>>

The required lateral sampling in an ultrasound image for a given center frequency, $f_c$, and bandwidth, $B$, is given by the aperture size, $D$, and the wavelength of the center frequency, $\lambda_c = c/f_c$, where $c$ is the speed of sound. The required Nyquist beam spacing is then \cite{Hergum2007}:
\begin{align}
\delta\theta < \frac{\lambda_c}{qpD}, \label{eq:resolution}
\end{align}
where the oversampling factor $p$ is applied to reduce folding of frequencies between $f_c$ and $f_c + B/2$. A $p$-factor of two has been used throughout this paper. The oversampling factor $q$ will be used to oversample prior to Capon beamforming and is one if not stated otherwise.

@@

\subsection{Oversampling by Phase Rotation}
The Capon beamformer formula in (\ref{eq:capon_weights}) gives a direct opportunity for oversampling using phase rotation. The steering vector $\vec{a}$ has in previous work been set to $\vec{1}$ since ultrasound data have to be pre-delayed. However, the steering vector can also, as in narrow band applications, be varied over a set of pre-defined vectors. In a narrow band setting the beam can make a sweep across the whole imaging sector through phase steering. This is also possible in broadband applications as long as the phase rotation is less than one pulse length (See coarse-fine beamforming in \cite{Thomenius} and its references). The maximal steering angle is approximately given by
\begin{align}
\theta_{max} = \alpha p \delta\theta,
\end{align}
where $\alpha$ is the pulse length in $\lambda$'s, and $p$ and $\delta\theta$ is the oversampling factor and beam spacing respectively as defined in (\ref{eq:resolution}). For our simulations $\theta_{max} = 3\delta\theta$.
It is therefore possible to make a sweep based on phase rotation between a given beam and half-way to its two nearest neighbours without introducing large errors. Phase rotation is actually what the Capon beamformer uses internally to micro-steer away from bright point scatterers and to enhance edges.

For a uniform linear array located along the x-axis the steering vector $\vec{a}_\theta$ should be calculated as 
\begin{align}
\vec{a}_\theta = 
\begin{bmatrix}
e^{-j\frac{2\pi}{\lambda_c}x_0\sin(\theta)} \\
e^{-j\frac{2\pi}{\lambda_c}x_1\sin(\theta)} \\
\vdots \\ 
e^{-j\frac{2\pi}{\lambda_c}x_{L-1}\sin(\theta)}
\end{bmatrix}
\end{align}
where $x_i$ is the element position and $\theta$ is swept from $-\delta\theta/2$ to $\delta\theta/2$. Note that the steering vector must be calculated for an $L$ element symmetric subarray, even though the full array has $M$ elements. This subarray should have the same element pitch, $x_i - x_{i-1}$, as the full array. %The steering error introduced by this steering vector for a frequency band $B$ is less than $B/2 \times \delta\theta/2$.

### 2014-02-23 ThesisJP/Background/background.tex
Changed symbol for resolution.

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end and a rasterized rendered image was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, the Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) instruction \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.1 GHz. This adds
\begin{equation}
16*40*1.1 = 704\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent. Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five or ten\footnote{Until an eight core Haswell CPU is launched later this year.} in raw computationally power. A claimed speed-up larger than this is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the pipeline. When GPGPU gained interest it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework called \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

No doubt, it is still a major overhead involved when writing GPU code, and in the future we will probably see work aimed at reducing this overhead.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions is handled in CPU code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding a heterogeneous core where some operations are automatically offloaded to special cores (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging.

@@

Sampling rates, and data size.
%Due to the high bandwidth requirement for medical ultrasound imaging, as we will see in Section \ref{sec:ultrasound}, a cloud computing ultrasound imaging system will not feasibly for a long time.
Because of the high data rate and processing requirements ultrasound systems where first analog systems up to the mid 90s. The first digital systems where made with custom boards and ASICs.

>>>>

Around the same time as CPU manufactures hit the frequency wall, people started experimenting with programmable shaders which recently had been introduced in the field of graphics programming. The graphics processing pipeline had before this been a fixed function pipeline, where e.g. geometry and textures where fed to the graphics processing unit (GPU) in one end, and where shaded a rasterized geometry was outputted in the other. What programmable shaders  introduced was the ability to replace some steps in the fixed-function pipeline with custom computations. This made it possible to utilize the GPU for other tasks than just rendering graphics \cite{Seland2007}. The rationale behind this exploration was that GPUs were already highly available in consumer PCs, and hidden behind its graphics interface were processing powers an order of magnitude larger than what the CPU could provide (Fig.\,\ref{fig:cpu_vs_gpu}). Actually, at the time when CPU manufacturers hit the frequency wall the GPU had already gone multi-core, first of all driven by an ever-increasing demand for more realistic computer games. Second, the rendering pipeline exhibits minimal execution path branching, and data are typically used once. Designers of GPUs could therefore skip advance feature, found in most CPUs, like branch prediction and different levels of data caching. These measures reduced the silicon footprint of each core and made it possible to add multiple cores to one GPU before CPU designers where able to do the same. Today, even if more advanced caching has been added to GPUs, this is still the main reason why GPUs have higher peak performance than CPUs (Fig.\,\ref{fig:cpu_vs_gpu}). Another reason is the inherent parallel nature of the rendering problem. When geometry is shaded, projected and rasterized the same instruction is typically needed for a lot of data at the same time. This makes a special kind of instruction, known as Single Instruction Multiple Data (\nom{SIMD}{Single instruction multiple data (instruction)}) \cite{Flynn1966}, especially suited for this kind of problem. The differences between how SIMD instructions are utilized in modern CPUs and GPUs are discussed in Section \ref{sec:cpu_vs_gpu}.

@@

In Section \ref{sec:gpgpu} we introduced a special type of instructions, namely SIMD instructions, which are designed to execute a single instruction across multiple data element in one clock cycle. At the end of the 20th century, Intel introduced a SIMD instruction set for their Pentium III processor called the Streaming SIMD Extensions (\nom{SSE}{Streaming SIMD extensions (instruction)}). This extension has seen several updates since. The first SEE instructions set had 128 bit long registers which meant that four, 32 bit, floating point values could be processed per clock cycle\footnote{For simplicity we will assume that the whole register is processed in one clock cycle even if the actual implementation might distributes one instruction for the register across multiple cycles.}. Updates of the SSE instruction set have added new instructions (SSE2, SSE3, and SSE4) and longer registers (\nom{AVX}{Advanced vector extensions (instruction)}, AVX2, and AVX-512). AVX and AVX2 adds 256 bit long registers and fused multiply-accumulate instructions (\nom{FMA}{Fused multiply-accumulate (instruction)})  which means that sixteen 32 bit floating point values can be processed per clock cycle per core. A new extension, AVX-512, extends the registers to 512 bit and is found in the Intel accelerator board  Xeon Phi and might show up in the next generation of CPUs from Intel. The latest CPU architecture from Intel, the Haswell architecture, has two processing ports supporting FMA-AVX2 instructions, and is therefore capable of processing 32, 32 bit, floating point values per clock cycle per core (Fig.\,\ref{fig:cpu_gpu}). For an eight core\footnote{An 8-core Haswell CPU is scheduled to launch during 2014.} CPU with a clock frequency of 3.8 GHz this sums to 
\begin{equation}
32*8*3.8 = 972.8\,\text{GFLOPS},
\end{equation}
hence a theoretical peak performance of close to one tera single precision floating point operations per second (TFLOPS). This is the same number as visualized in Fig\,\ref{fig:cpu_vs_gpu} for the Haswell architecture. In addition, the most powerful integrated GPU found on Haswell CPUs (HD 5200) have 40 execution units capable of processing sixteen, 32 bit, floating points at 1.3 GHz. This adds
\begin{equation}
16*40*1.3 = 832\,\text{GFLOPS}
\end{equation}
in extra processing power to the APU.

Modern GPUs can also be interpreted as a unit consisting of multiple SIMD processors (Fig.\,\ref{fig:cpu_gpu}). The Kepler architecture by Nvidia have up to 15 SIMD units capable of processing 192, 32 bit, floats per clock cycle at 745 MHz. These GPUs are also capable of performing FMA instructions, effectively doubling the number of theoretical FLOPS. The theoretical peak single precision FLOPS for this architecture is by that found to be 
\begin{equation}
15*192*2*745 = 4291\,\text{GFLOPS}.
\end{equation} 
Each SIMD core also comprise 16 special function units (\nom{SFU}{Special function unit (sin, exp, sqrt, ect.)}) which add close to 1 TFLOPS in additional computational power. This adds up to a total of five TFLOPS in computationally power as shown in Fig.\,\ref{fig:cpu_vs_gpu} for the GeForce GTX 780 TI GPU. From Fig.\,\ref{fig:cpu_gpu} we see how the high number of SIMD cores found on GPUs and their larger width, compared with SIMD cores on the CPU, results in five times the computational performance of a high-end CPU (without the integrated GPU).

Note that the derived numbers are strictly theoretical and that actual throughput is highly algorithm dependent. Nevertheless, in theory, the benefit of GPU computing over CPU computing is currently a factor of five in raw computationally power. A claimed speed-up of several order of magnitude is typically caused by comparing a non-optimized CPU implementation with an optimized GPU implementation or by comparing uneven hardware \cite{Lee2010, Kothapalli2013}. Also note that in this section we have focused on CPUs from Intel and GPUs from Nvidia, however, CPUs and GPUs from AMD have similar specifications.

@@

\subsection{Programming a GPU}
Early GPGPU programming required expert knowledge about computer graphics and the remaining fixed functions in the pipeline. When GPGPU gained interest it quickly became evident that new programing languages where needed.  One of the first GPGPU oriented languages for the GPU was the Brook language by Buck \textit{et al.} \cite{Buck2004}. Buck later joined Nvidia where he led the work on \nom{CUDA}{Compute unified device architecture (GPGPU language)}, a GPGPU language and framework by Nvidia for Nvidia GPUs only. The first version of CUDA was released in 2007. The following year, Apple launched a GPGPU framework known as \nom{OpenCL}{Open Compute Language (GPGPU language)}, a standard which are maintained by the Khronos Group. These two frameworks are still today the main programming frameworks for GPGPU. Where CUDA only runs on Nvidia GPUs, OpenCL can now run on GPUs from both Nvidia and AMD, CPUs from both Intel and AMD, FPGAs from Altera, and mobile processors from \nom{ARM}{Designer of mobile chip architectures}.

@@

No doubt, it is still a major overhead involved with writing GPU code, and in the future we will probably see work aimed at reducing this overhead.  A recent development is the introduction of C++ AMP which enables developers to write accelerated code (e.g. by a GPU) in plain C++. Hopefully, in the future, GPU programming will be a job for compilers, and direct fiddling with GPU kernels should only be needed if maximum throughput is absolutely needed. This is similar to how SIMD instructions is handled in CPU code today. Most of the time developers will rely on the CPU compilers to analyze the code and auto-apply SIMD instruction whenever it is possible. However, when maximum throughput is needed, developers with deep knowledge of the problem at hand are often able to make faster code by manually writing SIMD instruction code. The relevance of auto-applied GPU instructions by compilers will also increasing due to the recent embedding of GPUs on the same chip as the CPU. Soon, when writing code, we will probably talk about coding for a collection of heterogeneous cores where operations are automatically offloaded to special cores (e.g. a GPU) by the compiler.

\section {Medical ultrasound imaging}\label{sec:ultrasound}
Ultrasound imaging encompasses technology which generates images based on sound whose frequencies we can not hear. For medical ultrasound imaging frequencies between 2 MHz and 10 MHz are typically used. Frequency dependent attenuation force us to select a lower frequency if the object we want to image is located deep into the body. This is the reason why 2-4 MHz is used for cardiac imaging and a frequency around 7 MHz is used for vascular imaging. A piezoelectric ceramic is used to generate the ultrasonic frequencies, and is special in that way that ultrasound vibrations are generated when voltage is applied and \textit{vice versa}. It can therefore be used to both send and receive ultrasound signals, like a combined microphone and speaker. Piezoelectric elements organized in an array, is referred to as an ultrasound \textit{probe}, and a special probe is typically design for each medical imaging modality. For cardiac ultrasound imaging the probe footprint has to be smaller than the intercostal opening in order to avoid shadowing by ribs.

If the width of a piezoelectric element is less than the wavelength of the transmitted signal the element is said to be \textit{omnidirectional}, hence energy is distributed equally in all directions. When multiple omnidirectional elements are arranged in an array the transmitted signal will get shaped into a beam of ultrasound by diffraction. The 3 dB angular width of this beam, and also the angular resolution of the array, is given by \cite{AngelUltrasound}:
\begin{align}
\Delta = \frac{\lambda}{D},
\end{align}
\nomi{$\Delta$}{Angular resolution} where \nom{$\lambda$}{Wavelength} is the ultrasound wavelength and \nom{$D$}{Aperture width} is the aperture width. With a speed-of-sound in human tissue of $1540\frac{\text{m}}{\text{s}}$ the angular resolution of a 2 cm cardiac array at 3 MHz is around 1.5\degree, which at 8 cm gives a lateral resolution of 2 mm. The axial resolution is inverse proportional to the pulse length, hence the frequency bandwidth of the probe.

If time delays are applied on the signal sent to the different elements it is possible to control the steering and focusing of the beam. This technique is know as \textit{beamforming}, and for ultrasound imaging it is applied on both transmit and receive.

The standard form of time domain array beamforming is the delay-and-sum beamformer where the output $z$ is calculated as:
\begin{align}\label{eq:das}
z[n] = \sum_{m = 0}^{M-1}\vec{w}_m^*\vec{x}_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}[n],
\end{align}
where $M$ is the number of elements or channels in the array, $\Delta_m[n]$ is the per-element focusing and steering delay, and $\vec{w}$ is a weight vector or window function. The weight vector is typically selected in order to trade resolution for a lower side lobe level. Hence, the weights are typically real and their K-space response is then symmetric.

@@

Sampling rates, and data size.

### 2014-02-23 ThesisJP/Introduction/introduction.tex
Changed symbol for resolution.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

>>>>

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide more than 5 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.
